Isolator: accurate and consistent analysis of RNA-Seq experiments

### 2015-10-25 paper.tex
Figure tweaks.

\begin{table}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.884}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
\label{table:taqman}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.852}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 18353 genes using PrimePCR qPCR and RNA-Seq quantification.}
\label{table:primepcr}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.979} & \textbf{0.978} & \textbf{0.981} &
\textbf{0.982}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.}
\label{table:ercc}
\end{table}

@@

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.975} & \textbf{0.975}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between transcript-level estimates
for the mixed samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D.}
\label{table:seqctranscripts}
\end{table}

@@

\begin{table}\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & 0.919
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between ground truth and estimates
produced by each method on simulated RNA-Seq.}
\label{table:rlsim}
\end{table}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{rhemac-figure.pdf}
\caption{Proportionality correlation between estimates from 4.5 million 300nt
MiSeq reads and progressively larger numbers of HiSeq 2000 reads.}
\label{fig:rhemac}
\end{figure}

@@

Because the likelihood function in this problem is simple, and we sample
from the posterior probability directly, without introducing latent
variables, we are afforded some optimization opportunities not available
in a Gibbs sampler. Specifically, we implement the likelihood function
using SIMD (single instruction, multiple data) instructions, which are
available nearly all CPUs from the last decade. This, in combination
with some numerical approximations, allow us to compute the likelihood
function over an order of magnitude faster on conventional hardware,
than a naive C implementation.

TODO: point to timing table in supplement

>>>>

\begin{table}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.883}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
\label{table:taqman}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.851}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.}
\label{table:primepcr}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.980} & \textbf{0.979} & \textbf{0.981} &
\textbf{0.983}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.}
\label{table:ercc}
\end{table}

@@

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.978} & \textbf{0.978}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between transcript-level estimates
for the mixed SEQC samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D. Sample C consists of
$3/4$ A and $1/4$ B and sample D is $1/4$ A and $3/4$ B. Consistent estimates of
transcript expression in A, B, C, and D should agree with these mixture
proportions.}
\label{table:seqctranscripts}
\end{table}

@@

\begin{table}\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & 0.919
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between ground truth and estimates
produced by each method using simulated RNA-Seq reads.}
\label{table:rlsim}
\end{table}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{rhemac-figure.pdf}
\caption{Proportionality correlation between estimates from 4.5 million 300nt
MiSeq reads and progressively larger numbers of HiSeq 2000 reads. For clarity,
correlations for RSEM/PM are excluded as they were much lower than the other
methods.}
\label{fig:rhemac}
\end{figure}

@@

Because the likelihood function in this problem is simple, and we sample
from the posterior probability directly, without introducing latent
variables, we are afforded some optimization opportunities not available
in a Gibbs sampler. Specifically, we implement the likelihood function
using SIMD (single instruction, multiple data) instructions, which are
available nearly all CPUs from the last decade. This, in combination
with some numerical approximations, allow us to compute the likelihood
function over an order of magnitude faster on conventional hardware,
than a more straightforward implementation. Run time for Isolator is generally
significantly less than that of other sampling based implementations, though
alignment-free, maximum-likelihood methods like Salmon and Kallisto remain the
most efficient methods (Supplementary Figure \ref{fig:runtime})

### 2015-10-25 supplement.tex
Stuff on qpcr validation.
Figure tweaks.

% TODO:
%   * Explain that every SEQC sample was put in the same experiment for
%     isolator.
%   * Move gene-expression consistency table here.
%   *

@@

\begin{table}
\begin{tabular}[c]{ll}
    STAR & 2.4.0k \\
    Cufflinks & 2.2.1 \\
    BitSeq & 0.7.0 \\
    RSEM & 1.2.19 \\
    Sailfish & 0.6.3 \\
    eXpress & 1.5.1 \\
    Kallisto & 0.42 \\
    Salmon & 0.4.1
\end{tabular}
\caption{
    Version of software used in the analysis.
}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D. TODO: move this
table to the supplement.}
\label{table:seqcgenes}
\end{table}

@@

\subsection{qPCR Validation}

Total RNAs for 2 independent HFA and HFV samples were extracted using RNAeasy
mini kit (Qiagen) (6). Two independent HAH total RNA samples were purchased
commercially from Clontech and Agilent. Total RNAs for 3 independent replicates
of Day 20 cardiomyocytes (derived from human embryonic stem cells) were
extracted with miRNeasy mini kit (Qiagen). All RNA samples were further DNAse
treated with DNA-free kit (Ambion) and reverse transcription reaction was
carried out using Omniscript RT kit (Qiagen). Gene specific primers were
designed targeting the exon-exon junction for specific exons of interest using
the freely available software APE. The primer sequences were as follows:
TNNT2-F- CAGAGACCATGTCTGACATAGA and R-CTGCCTCCTCCTGCTCGT; OBSL1-F-
CCTGTACGCTGGTACAAGGA and R-CTCTTCCACGCTGACAAGGA; LTBP4-F- CAGCCCAAAAAGTGTGCAGG
and R-CTGTCTCTTCCTCACGCTGG; OSBPL1A-F- TCACAGCTCTGCTCAACAGG and
R-ACATTGTTTATGGGCCCGGT; ACOT9-F-CCACATTCATGAAGCATGTTCA and
R-CCTACTATCTCCCGCAACT, IMMT-F-TGCACTTTCAGAAGAAGCATCC and R-TTTTCCTGTTGTGCAAGGCG;
PALM-F-TGAAGGCAGCCATGTACTCG and R-GGACCACTTTGGTCTCGTCC

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\end{figure}

\subsection{Principal Component Analysis}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/isolator-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/salmon-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/sailfish-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/kallisto-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/express-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/cufflinks-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/pm-rsem-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ml-rsem-pca.pdf}
\caption{The first two principal components from principal component analysis
applied to transcript expression estimates using a variety of methods.}
\end{figure}

>>>>

% TODO:
%   * Explain that every SEQC sample was put in the same experiment for
%     isolator.

@@

\begin{table}
\begin{center}
\begin{tabular}[c]{ll}
\toprule\addlinespace
Method & Version \\
\midrule
STAR & 2.4.0k \\\addlinespace
Cufflinks & 2.2.1 \\\addlinespace
BitSeq & 0.7.0 \\\addlinespace
RSEM & 1.2.19 \\\addlinespace
Sailfish & 0.6.3 \\\addlinespace
eXpress & 1.5.1 \\\addlinespace
Kallisto & 0.42 \\\addlinespace
Salmon & 0.4.1 \\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Versions of software used in the analysis.}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D. TODO: move this
table to the supplement.}
\end{center}
\label{table:seqcgenes}
\end{table}

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{analysis/seqc/runtime.pdf}
\caption{Run time needed to process the SEQC data presented in the Results
sections. All methods were run a Google Compute Engine instance backed by four
Intel Xeon cores and 52GB of a memory.}
\end{center}
\label{fig:runtime}
\end{figure}

@@

\subsection{qPCR Validation}\label{cardioqpcr}

Total RNAs for 2 independent HFA and HFV samples were extracted using RNAeasy
mini kit (Qiagen) (6). Two independent HAH total RNA samples were purchased
commercially from Clontech and Agilent. Total RNAs for 3 independent replicates
of Day 20 cardiomyocytes (derived from human embryonic stem cells) were
extracted with miRNeasy mini kit (Qiagen). All RNA samples were further DNAse
treated with DNA-free kit (Ambion) and reverse transcription reaction was
carried out using Omniscript RT kit (Qiagen). Gene specific primers were
designed targeting the exon-exon junction for specific exons of interest using
the freely available software APE. The primer sequences were as follows:

\begin{center}
\begin{tabular}{rl}
TNNT2   & F-CAGAGACCATGTCTGACATAGA \\
        & R-CTGCCTCCTCCTGCTCGTx \\
OBSL1   & F-CCTGTACGCTGGTACAAGGA \\
        & R-CTCTTCCACGCTGACAAGGA \\
LTBP4   & F-CAGCCCAAAAAGTGTGCAGG \\
        & R-CTGTCTCTTCCTCACGCTGG \\
OSBPL1A & F-TCACAGCTCTGCTCAACAGG \\
        & R-ACATTGTTTATGGGCCCGGT \\
ACOT9   & F-CCACATTCATGAAGCATGTTCA \\
        & R-CCTACTATCTCCCGCAACT \\
IMMT    & F-TGCACTTTCAGAAGAAGCATCC \\
        & R-TTTTCCTGTTGTGCAAGGCG \\
PALM    & F-TGAAGGCAGCCATGTACTCG \\
        & R-GGACCACTTTGGTCTCGTCC
\end{tabular}
\end{center}

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\label{fig:cardiofoldchange}
\end{figure}

\subsection{Principal Component Analysis}\label{cardiopca}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/isolator-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/salmon-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/sailfish-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/kallisto-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/express-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/cufflinks-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/pm-rsem-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ml-rsem-pca.pdf}
\caption{The first two principal components from principal component analysis
applied to transcript expression estimates using a variety of methods.}
\label{fig:cardiopca}
\end{figure}
